Computing likelihoods under the SMC for a general class of ARGs.

### 2023-11-16 paper.tex
update recording ARGs

%\usepackage{subcaption}

\newcommand{\noderef}[1]{\textsf{#1}}
\newcommand{\tsinfer}[0]{\texttt{tsinfer}}
\newcommand{\kwarg}[0]{\texttt{KwARG}}
\newcommand{\argweaver}[0]{\texttt{ARGweaver}}
\newcommand{\argweaverD}[0]{\texttt{ARGweaver-D}}
\newcommand{\relate}[0]{\texttt{Relate}}
\newcommand{\espalier}[0]{\texttt{Espalier}}
\newcommand{\arbores}[0]{\texttt{Arbores}}

@@

@@

@@

@@

% then onto the SMC
[TO ADD: discuss SMC] mention the SMC \citep{mcvean_approximating_2004}.
SMC and relatives are useful: lot of focus however on integrating over 
the distribution of trees to get $N_e(t)$ [ADD REF: PSMC, MSMC]
Here, the idea is to use the SMC in a different way that resonates more 
with the true direction of flow of genetic information.
Results in a concrete way to compute likelihoods for *any* inferred ARG 
by integrating out the time to recombination.
and formulate an algorithm 
that is general in the sense that it can deal with the various levels of 
completeness that characterise the field's currents attempts [TO DO: make this sound more positive] 
at reconstucting genome-wide genealogies.

@@

More formally, and following the notation of \citet{mcvean_approximating_2005}, at any 
point in time, the process can be described by $L(t)$, the set of labeled lineages 
extant at time $t$ each represented by a union of ordered non-overlapping half open 
intervals detailing the ancestral material 
carried by that lineage $X_i = \{[x_{i0}, y_{i0}), \dotsc, [x_{im}, y_{im})\}$.
$L(0)$ consists of $n$ sampled lineages, labelled $0$ to $n-1$, each represented by a 
single interval spanning the entire genome.
Backwards in time the process evolves by a succession of coalescence and recombination 
events until all segments of ancestral material are each present in only one lineage. 
The waiting time to next event is determined by these two competing processes with exponentially 
distributed waiting times as outlined below.\\

@@

The process described in the previous paragraph details the transitions of the 
set of labeled lineages $L$ through time. An Ancestral Recombination Graph 
$\mathcal{G} = (N, E)$ capturing the various inheritance paths along the genome 
is obtained by registering a tuple $(c, p, X_{cp})$ whenever a 
lineage $c$, associated with a disjoint set of genomic intervals $X_c$, is 
replaced by one or two lineage(s), $p$ (and $p^{\prime}$), 
in case of a coalescence or recombination respectively.
The last element of the tuple is the extent of the edge. It is itself a  
disjoint set of intervals such that $X_{cp} \cup X_{cp^{\prime}} = X_c$. 
The set of edges $E$, containing all such tuples $(c, p, X_{cp})$, 
and $N$, the set of nodes representing all genomes in the ARG, together
define this data structure.\\

Recording the full process as described above will result in unary nodes.
These are nodes that have only one child in one or more local trees 
(see fig.\ref{fig:smc-unary} and \citet{Wong-2023}). This is not only true 
for the nodes registered for those lineages affected by a recombination event. 
Coalescence between two lineages without perfectly 
overlapping intervals will similarly result in locally 
unary nodes (see \ref{fig:smc-unary}).These unary nodes ensure the one 
to one correspondence between an ARG and its 
representation as a sequence of local trees. 
Indeed, the presence of these unary nodes allow us to completely reconstruct 
the state of all extant lineages through time from the local trees.\\

%under the SMC up to the exact time of the recombination events.

The resulting data structure can be simplified to various degrees \citep{Wong-2023}. 
This equates to removing certain types of unary nodes, 
depending on the degree of information 
we want to retain (or can reconstruct) from the original process. 
In a fully simplified ARG, without unary nodes, 
the local trees only store information on local coalescence events.
Under the SMC, if we only omit recombination nodes, % is there a better word?
we are left with the coalescent nodes, those nodes that are coalescent 
along some fraction of the genome. We can however still identify \emph{all} 
lineages that were hit by a recombination event and constrain the time to 
this event to within a window defined by the age of the node $c$ associated with that 
lineage and the minimum age of the parents of the edges connecting to $c$.

As highlighted in fig.\ref{fig:smc-unary}, (coalescing) unary nodes can be inferred 
from the available haplotype information. Currently, the following methods 
return the full this information by default as part of their ARG reconstruction algorithm: 
\argweaver \citep{rasmussen_genome-wide_2014} 
\tsinfer \citep{kelleher_inferring_2019}, 
\kwarg \citep{ignatieva_kwarg_2021}.

The presence of these (coalescing) unary nodes enables us to extend likelihood 
computations to a more general class of ARGs.
Here we will compute the likelihood for an ARG
under the SMC for which we have no information on the time to any 
of the recombination events. 
We will assume however, that the identity 
of the lineage involved in each recombination event can be correctly 
inferred by relying on the unary nodes.

@@

%\subsection{Unary nodes} \label{par:unary}
% Do we need a seperate section on the unary nodes?
% is there more to be said about the SMC inm the presence of unary nodes?
%Recording unary nodes has interesting consequences on the left-to-right 
%description of the SMC. Looking at a single marginal tree, the presence of 
%unary nodes are telltale signs of future or past coalescence events (in a 
%left-to-right sense).\\

@@

Because the likelihood is computed based on edges,
and edges are no longer recorded once the local most recent common ancestor 
has been reached, we will underestimate the number of positions where 
a recombination could have occurred in case a recombination hits a 
lineage with trapped non-ancestral material (see end \ref{par:description}). 
If we assume that only a single recombination could have  
occurred, then it suffices however to provide a correction factor $g$ in equation
\ref{eq:depth}. Here $g$ represents the 
length of non-ancestral material along which a recombination would have resulted  
in the same observed local sequence of tree changes.

@@

% polytomies
[TO DO: polytomies] The above algorithm treats each edge independently which implies 
we can deal with polytomies.

@@

Here, we have introduced a scalable algorithm to compute likelihoods 
under the SMC for a general class of ARGs. By formulating the SMC backwards in time 
we were able to associate a single likelihood value with each edge in the ARG. 
The compatibility of the algorithm with the \tskit library allows us to 
exploit its associated efficiencies both for the current work and its future 
applications. This work will contribute significantly 
to ARG-based analysis becoming a standard part of the population 
and statistical genetics toolkit. 
In particular, we see three main applications for the work presented here. 
Firstly, the ability to associate inferred ARGs with a likelihood given a demographic 
model and an associated set of parameters will help bridge the gap between the tremendous recent 
advances in ARG inference and the current state of the art statistical inference of 
past demography. Secondly, hybrid ARG-inference methods that combine the best of both 
heuristic and model-based inference approaches are now possible.
Thirdly, using a model we can quantify the uncertainty inherently associated 
with ARG-reconstruction.

>>>>

\newcommand{\tsinfer}[0]{\texttt{tsinfer}}
\newcommand{\kwarg}[0]{\texttt{KwARG}}
\newcommand{\argweaver}[0]{\texttt{ARGweaver}}
\newcommand{\argweaverD}[0]{\texttt{ARGweaver-D}}
\newcommand{\relate}[0]{\texttt{Relate}}
\newcommand{\espalier}[0]{\texttt{Espalier}}
\newcommand{\arbores}[0]{\texttt{Arbores}}
\newcommand{\tskit}[0]{\texttt{tskit}}

@@

[great point from Wong et al] Note that it is important to distinguish 
here between structures and models: whether an inference method is based on 
heuristics or a rigorous mathematical model is orthogonal to the level 
of detail provided in its estimate. One could heuristically estimate a 
fully precise ARG, or statistically sample a partial, approximate ARG under 
a model such as the coalescent.
What we want to achieve with this paper is to bring the two concepts closer 
to each other. Compute a likelihood for an approximate (define) ARG under a 
approximate model of the coalescent. 
Approximate in two senses: SMC + leave out precise recombination time info

@@

The breakthroughs in scale achieved by recent methods 
are all based, in different ways, on inferring approximate 
structures instead of a complete and fully detailed history
% in particular with respect to recombination nodes

@@

[TO DO:] what does *general* mean? and when in the paper do we discuss it?

@@

% then onto the SMC
[TO ADD: discuss SMC] mention the SMC \citep{mcvean_approximating_2004}.
SMC and relatives are useful: lot of focus however on integrating over 
the distribution of trees to get $N_e(t)$ [ADD REF: PSMC, MSMC]
Here, the idea is to use the SMC in a different way that resonates more 
with the true direction of flow of genetic information.

Results in a concrete way to compute likelihoods for *any* inferred ARG 
by integrating out the time to recombination.
and formulate an algorithm 
that is general in the sense that it can deal with the various levels of 
completeness that characterise the field's currents attempts [TO DO: make this sound more positive] 
at reconstucting genome-wide genealogies.

@@

More formally, and following the notation of \citet{mcvean_approximating_2005}, at any 
point in time, the process can be described by $L(t)$, the set of labeled lineages 
extant at time $t$ each represented by a union of ordered non-overlapping half open 
intervals detailing the ancestral material 
carried by that lineage $X_i = \{[x_{i0}, y_{i0}), \dotsc, [x_{im}, y_{im})\}$.
$L(0)$ consists of $n$ sampled lineages, labelled $0$ to $n-1$, each represented by a 
single interval spanning the entire genome.
Backwards in time the process evolves by a succession of coalescence and recombination 
events until each segment of ancestral material is only present in one lineage. 
The waiting time to next event is determined by these two competing processes 
with exponentially distributed waiting times as outlined below.\\

@@

The process described in the previous paragraph details the transitions of the 
set of labeled lineages $L$ through time. An Ancestral Recombination Graph 
$\mathcal{G} = (N, E)$ capturing the various inheritance paths along the genome 
is obtained by registering an edge represented by a tuple $(c, p, X_{cp})$ 
whenever a lineage $c$, associated with a disjoint set of genomic intervals $X_c$, is 
replaced by one or two lineage(s), $p$ (and $p^{\prime}$), 
in case of a coalescence or recombination respectively.
The last element of the tuple is the extent of the edge. It is itself a  
disjoint set of intervals such that $X_{cp} \cup X_{cp^{\prime}} = X_c$. 
The set of edges $E$, containing all such tuples $(c, p, X_{cp})$, 
and $N$, the set of nodes representing all genomes in the ARG, together
define this data structure.\\

The ARG as described here records the impact of every possible event along each 
inheritance path in full detail. We could however simplify this graph by 
throwing away the precise time information of all recombination events. This 
corresponds to removing the recombinant internal nodes within the ARG such that some of 
the internal nodes representing coalescence events now have 2 outgoing (backwards in time) 
edges. On the ARG in fig.\ref{fig:smc-unary}, the output of a simulation under the SMC, 
this corresponds to removing the blue internal nodes. Crucially however, 
we can still identify all lineages that were hit by a recombination event. Yet we 
can only constrain the time to this event to within a window defined by the age of the 
child node $c$ associated with that lineage and the minimum age of the parents of the 
edges connecting to $c$.\\

ARG simplification therefore touches upon the question of what is knowable.  
[TO DO: Add link with inference methods]\\

Many ARG reconstruction methods return a data structure that 
is in essence a series of local tries [REFs]. This does not need to be a contradiction. 
All ARGs can be represented as a series of local trees \citep{wong-2023}.
As long as we register each observed node when building the local tree that 
records the path through the ARG, then we can guarantee a one-to-one correspondence 
with the original ARG. Crucially, this will result in nodes that 
are locally unary. These are nodes that only have a single child along some of the 
intervals within the genome.\\

Recombination nodes are one such type of unary nodes. Figure \ref{fig:smc-unary} (bottom)
reveals however that all internal nodes but node 4 are unary along some part of 
the genome. These nodes all represent coalescence events.

Note that the presence of these unary nodes allows us to reconstruct the full extent 
of each lineage going back through time. Similarly, all lineages involved in a recombination 
event can be uniquely identified by a change in parent going from one tree to the next 
for the child node associated with that lineage. Both the ARG and the sequence of 
local trees contain the same information, as long as they are simplified to the 
same degree.\\

[TO DO: do I need to expand on this?] Simplification is thus tied in with the concept of locally unary nodes.\\

Which ARG inference methods return locally unary nodes and how can they be inferred.
[TO DO] give a good explanation (using the figure) why these methods can infer the extent 
of coalescing lineages.
As highlighted in fig.\ref{fig:smc-unary}, (coalescing) unary nodes can be inferred 
from the available haplotype information. Currently, the following methods 
return the full this information by default as part of their ARG reconstruction algorithm: 
\argweaver \citep{rasmussen_genome-wide_2014} 
\tsinfer \citep{kelleher_inferring_2019}, 
\kwarg \citep{ignatieva_kwarg_2021}.

[TO DO]: impact of unary nodes on reasoning about the SMC in a left-to-right sense.

In the next paragraph we will derive an expression to compute the 
likelihood for an ARG under the SMC for which we have no information on the 
time to any of the recombination events. We will then detail an algorithm 
that allows us to compute this value given an ARG 
represented as a series of local trees. To do this we will rely on the unary nodes 
to identify each lineage involved in a recombination event 
as described in this section.

@@

@@

[TO DO: improve this remark: when recording an ARG we do record edges in this situation,
however, msprime simulations will stop recording edges for local trees that have reached 
their local mrca.] 
Because the likelihood is computed based on edges,
and edges are no longer recorded once the local most recent common ancestor 
has been reached, we will underestimate the number of positions where 
a recombination could have occurred in case a recombination hits a 
lineage with trapped non-ancestral material (see end \ref{par:description}). 
If we assume that only a single recombination could have  
occurred, then it suffices however to provide a correction factor $g$ in equation
\ref{eq:depth}. Here $g$ represents the 
length of non-ancestral material along which a recombination would have resulted  
in the same observed local sequence of tree changes.\\

@@

% polytomies
[TO DO: polytomies] The above algorithm treats each edge independently which implies 
we can deal with polytomies.
[TO DO: discuss, what if the identity of internal nodes is not preserved across trees]

@@

Here, we have introduced a scalable algorithm to compute likelihoods 
under the SMC for a general class of ARGs. By formulating the SMC backwards in time 
we were able to associate a single likelihood value with each edge in the ARG. 
The compatibility of the algorithm with the \texttt{tskit} library allows us to 
exploit its associated efficiencies both for the current work and its future 
applications. This work will contribute significantly 
to ARG-based analysis becoming a standard part of the population 
and statistical genetics toolkit. 
In particular, we see three main applications for the work presented here. 
Firstly, the ability to associate inferred ARGs with a likelihood given a demographic 
model and an associated set of parameters will help bridge the gap between the tremendous recent 
advances in ARG inference and the current state of the art statistical inference of 
past demography. Secondly, hybrid ARG-inference methods that combine the best of both 
heuristic and model-based inference approaches are now possible.
Thirdly, using a model we can quantify the uncertainty inherently associated 
with ARG-reconstruction.
